---
title: "MachineLearning - Script 5 - Variation Partitioning"
author: 
  - name: "MSc. Friederike Johanna Rosa Wölke"
    orcid: "0000-0001-9034-4883"
    url: "https://friedarosa.github.io"
    email: "wolke@fzp.czu.cz"
    corresponding: true
date: "2023-05-29"
format: 
  html:
    toc: true
    code-fold: true
    code-overflow: wrap
    bibliography: references.bib
keep-tex: true
---

### Variation Partitioning between Hypotheses

Now we will look which of the four hypotheses explains most variation in the response.
First will will use the varpart function from the vegan package for this, and second we will do the same manually with random forest models.

::: panel-tabset
## Jaccard - Sampling Period 1

```{r}
#| label: var-part-vegan-J1
dat_train <- dat_train_J1
# Create a data frame for each set of variables
H1_data <- dat_train %>% select(any_of(H1_vars))
H2_data <- dat_train %>% select(any_of(H2_vars))
H3_data <- dat_train %>% select(any_of(H3_vars))
H4_data <- dat_train %>% select(any_of(H4_vars))

# Perform basic variation partitioning (vegan package)
varpart_model <- varpart(dat_train$Jaccard, H1_data, H2_data, H3_data, H4_data)

# Print the results
print(varpart_model)
summary(varpart_model)



## VENN-diagram
plot(varpart_model,
     Xnames = c("Species Traits", "Species Range Geometry", "Diversity", "Atlas characteristics"), # name the partitions
     bg = c("seagreen3", "mediumpurple", "darkorange", "gold"), alpha = 80, # colour the circles
     digits = 2, # only show 2 digits
     cex = 0.8,
     id.size = 1)


# Test for significance in results:
anova.cca(rda(dat_train$Jaccard, H1_data)) # sign. p = 0.001
anova.cca(rda(dat_train$Jaccard, H2_data)) # sign. p = 0.001
anova.cca(rda(dat_train$Jaccard, H3_data)) # sign. p = 0.001
anova.cca(rda(dat_train$Jaccard, H4_data)) # sign. p = 0.001

```

## Jaccard - Sampling Period 2

```{r}
#| label: var-part-vegan-J2
dat_train <- dat_train_J2
# Create a data frame for each set of variables
H1_data <- dat_train %>% select(any_of(H1_vars))
H2_data <- dat_train %>% select(any_of(H2_vars))
H3_data <- dat_train %>% select(any_of(H3_vars))
H4_data <- dat_train %>% select(any_of(H4_vars))

# Perform basic variation partitioning (vegan package)
varpart_model <- varpart(dat_train$Jaccard, H1_data, H2_data, H3_data, H4_data)

# Print the results
print(varpart_model)
summary(varpart_model)

## VENN-diagram
plot(varpart_model,
     Xnames = c("Species Traits", "Species Range Geometry", "Diversity", "Atlas characteristics"), # name the partitions
     bg = c("seagreen3", "mediumpurple", "darkorange", "gold"), alpha = 80, # colour the circles
     digits = 2, # only show 2 digits
     cex = 0.8,
     id.size = 1)

# Test for significance in results:
anova.cca(rda(dat_train$Jaccard, H1_data)) # sign. p = 0.001
anova.cca(rda(dat_train$Jaccard, H2_data)) # sign. p = 0.001
anova.cca(rda(dat_train$Jaccard, H3_data)) # sign. p = 0.001
anova.cca(rda(dat_train$Jaccard, H4_data)) # sign. p = 0.001
```

## Log Ratio - Sampling Period 1

```{r}
#| label: var-part-vegan-J1
dat_train <- dat_train_LR1
# Create a data frame for each set of variables
H1_data <- dat_train %>% select(any_of(H1_vars))
H2_data <- dat_train %>% select(any_of(H2_vars))
H3_data <- dat_train %>% select(any_of(H3_vars))
H4_data <- dat_train %>% select(any_of(H4_vars))

# Perform basic variation partitioning (vegan package)
varpart_model <- varpart(dat_train$log_R2_1, H1_data, H2_data, H3_data, H4_data)

# Print the results
print(varpart_model)
summary(varpart_model)



## VENN-diagram
plot(varpart_model,
     Xnames = c("Species Traits", "Species Range Geometry", "Diversity", "Atlas characteristics"), # name the partitions
     bg = c("seagreen3", "mediumpurple", "darkorange", "gold"), alpha = 80, # colour the circles
     digits = 2, # only show 2 digits
     cex = 0.8,
     id.size = 1)


# Test for significance in results:
anova.cca(rda(dat_train$log_R2_1, H1_data)) # sign. p = 0.001
anova.cca(rda(dat_train$log_R2_1, H2_data)) # sign. p = 0.001
anova.cca(rda(dat_train$log_R2_1, H3_data)) # sign. p = 0.001
anova.cca(rda(dat_train$log_R2_1, H4_data)) # sign. p = 0.001

```

## Log Ratio - Sampling Period 2

```{r}
#| label: var-part-vegan-J1
dat_train <- dat_train_LR2
# Create a data frame for each set of variables
H1_data <- dat_train %>% select(any_of(H1_vars))
H2_data <- dat_train %>% select(any_of(H2_vars))
H3_data <- dat_train %>% select(any_of(H3_vars))
H4_data <- dat_train %>% select(any_of(H4_vars))

# Perform basic variation partitioning (vegan package)
varpart_model <- varpart(dat_train$log_R2_1, H1_data, H2_data, H3_data, H4_data)

# Print the results
print(varpart_model)
summary(varpart_model)



## VENN-diagram
plot(varpart_model,
     Xnames = c("Species Traits", "Species Range Geometry", "Diversity", "Atlas characteristics"), # name the partitions
     bg = c("seagreen3", "mediumpurple", "darkorange", "gold"), alpha = 80, # colour the circles
     digits = 2, # only show 2 digits
     cex = 0.8,
     id.size = 1)


# Test for significance in results:
anova.cca(rda(dat_train$log_R2_1, H1_data)) # sign. p = 0.001
anova.cca(rda(dat_train$log_R2_1, H2_data)) # sign. p = 0.001
anova.cca(rda(dat_train$log_R2_1, H3_data)) # sign. p = 0.001
anova.cca(rda(dat_train$log_R2_1, H4_data)) # sign. p = 0.001

```
:::

::: panel-tabset
## Jaccard - Sampling Period 1

```{r}
#| label: var-part-rf
#| eval: FALSE
dat_train <- dat_train_J1

trainControl <- trainControl(
    method = "repeatedcv",
    number = 10,
    repeats = 3,
    savePredictions = "final",
    returnResamp = "final",
    verboseIter = TRUE,
    index = indices)

tictoc::tic("ranger full model")
set.seed(42)
full_ranger <- train(
    Jaccard ~ .,
    data = dat_train,
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


# Train ranger model
tictoc::tic("ranger H1")
set.seed(42)
H1_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(H1_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


tictoc::tic("ranger H2")
set.seed(42)
H2_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(H2_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


tictoc::tic("ranger H3")
set.seed(42)
H3_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(H3_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()

tictoc::tic("ranger H4")
set.seed(42)
H4_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(H4_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


### combinations of 2 hypotheses:

set.seed(42)
H1H2_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H3_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H4_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)


set.seed(42)
H2H3_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H2_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)


set.seed(42)
H2H4_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H2_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H3H4_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H3_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)


### combinations of 3 hypotheses together =====

set.seed(42)
H1H2H3_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H2H4_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H3H4_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H3_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H2H3H4_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

J1_varPart_ranger_list <- list(
    full_ranger, H1_ranger, H2_ranger, H3_ranger, H4_ranger, 
    H1H2_ranger, H1H3_ranger, H1H4_ranger, 
    H2H3_ranger, H2H4_ranger, 
    H3H4_ranger, 
    H1H2H3_ranger, H1H2H4_ranger, H1H3H4_ranger, 
    H2H3H4_ranger)
saveRDS(J1_varPart_ranger_list, "../data/J1_varPart_list.rds")

```

```{r}
#| label: var-part-ranger-performance-eval
# ===== Performance eval ======= #

# Predict on your test data
predictions_full <- predict(full_ranger, newdata = dat_test)

predictions_H1 <- predict(H1_ranger, newdata = dat_test)
predictions_H2 <- predict(H2_ranger, newdata = dat_test)
predictions_H3 <- predict(H3_ranger, newdata = dat_test)
predictions_H4 <- predict(H4_ranger, newdata = dat_test)

predictions_H1H2 <- predict(H1H2_ranger, newdata = dat_test)
predictions_H1H3 <- predict(H1H3_ranger, newdata = dat_test)
predictions_H1H4 <- predict(H1H4_ranger, newdata = dat_test)
predictions_H2H3 <- predict(H2H3_ranger, newdata = dat_test)
predictions_H2H4 <- predict(H2H4_ranger, newdata = dat_test)
predictions_H3H4 <- predict(H3H4_ranger, newdata = dat_test)

predictions_H1H2H3 <- predict(H1H2H3_ranger, newdata = dat_test)
predictions_H1H2H4 <- predict(H1H2H4_ranger, newdata = dat_test)
predictions_H1H3H4 <- predict(H1H3H4_ranger, newdata = dat_test)
predictions_H2H3H4 <- predict(H2H3H4_ranger, newdata = dat_test)

# Calculate the performance metrics
perf <- rbind(postResample(pred = predictions_full, obs = dat_test$Jaccard),
              postResample(pred = predictions_H1, obs = dat_test$Jaccard),
              postResample(pred = predictions_H2, obs = dat_test$Jaccard),
              postResample(pred = predictions_H3, obs = dat_test$Jaccard),
              postResample(pred = predictions_H4, obs = dat_test$Jaccard),
              
              postResample(pred = predictions_H1H2, obs = dat_test$Jaccard),
              postResample(pred = predictions_H1H3, obs = dat_test$Jaccard),
              postResample(pred = predictions_H1H4, obs = dat_test$Jaccard),
              postResample(pred = predictions_H2H3, obs = dat_test$Jaccard),
              postResample(pred = predictions_H2H4, obs = dat_test$Jaccard),
              postResample(pred = predictions_H3H4, obs = dat_test$Jaccard),
              
              postResample(pred = predictions_H1H2H3, obs = dat_test$Jaccard),
              postResample(pred = predictions_H1H2H4, obs = dat_test$Jaccard),
              postResample(pred = predictions_H1H3H4, obs = dat_test$Jaccard),
              postResample(pred = predictions_H2H3H4, obs = dat_test$Jaccard)) %>% 
                as.data.frame() %>% 
                round(4)

model <- c("full", 
            seq(1:4), 
            "H1H2", "H1H3", "H1H4", "H2H3", "H2H4", "H3H4", 
            "H1H2H3", "H1H2H4", "H1H3H4", "H2H3H4")
perf$model <- model

# Print the performance metrics

perf %>% 
    kableExtra::kable()
perf %>% 
    kableExtra::kable() %>% 
    write.csv("../data/performance_varExpl_rf_J1.csv")

slice_min(perf, RMSE) %>% 
    slice_max(Rsquared)


# Create a bar plot of variance explained
ggplot(perf, aes(x = reorder(model, RMSE), y = Rsquared)) +
  geom_bar(stat = "identity") +
  labs(x = "Model", y = "Variance Explained") +
  theme_minimal()

```

## Jaccard - Sampling period 2

```{r}
#| label: var-part-rf-J2
#| eval: FALSE
dat_train <- dat_train_J2

trainControl <- trainControl(
    method = "repeatedcv",
    number = 10,
    repeats = 3,
    savePredictions = "final",
    returnResamp = "final",
    verboseIter = TRUE,
    index = indices)

tictoc::tic("ranger full model")
set.seed(42)
full_ranger <- train(
    Jaccard ~ .,
    data = dat_train,
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


# Train ranger model
tictoc::tic("ranger H1")
set.seed(42)
H1_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(H1_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


tictoc::tic("ranger H2")
set.seed(42)
H2_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(H2_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


tictoc::tic("ranger H3")
set.seed(42)
H3_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(H3_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()

tictoc::tic("ranger H4")
set.seed(42)
H4_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(H4_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


### combinations of 2 hypotheses:

set.seed(42)
H1H2_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H3_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H4_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)


set.seed(42)
H2H3_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H2_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)


set.seed(42)
H2H4_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H2_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H3H4_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H3_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)


### combinations of 3 hypotheses together =====

set.seed(42)
H1H2H3_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H2H4_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H3H4_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H3_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H2H3H4_ranger <- train(
    Jaccard ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

J2_varPart_ranger_list <- list(
    full_ranger, H1_ranger, H2_ranger, H3_ranger, H4_ranger, 
    H1H2_ranger, H1H3_ranger, H1H4_ranger, 
    H2H3_ranger, H2H4_ranger, 
    H3H4_ranger, 
    H1H2H3_ranger, H1H2H4_ranger, H1H3H4_ranger, 
    H2H3H4_ranger)
saveRDS(J2_varPart_ranger_list, "../data/J2_varPart_list.rds")

```

```{r}
#| label: var-part-ranger-performance-eval
# ===== Performance eval ======= #

# Predict on your test data
predictions_full <- predict(full_ranger, newdata = dat_test)

predictions_H1 <- predict(H1_ranger, newdata = dat_test)
predictions_H2 <- predict(H2_ranger, newdata = dat_test)
predictions_H3 <- predict(H3_ranger, newdata = dat_test)
predictions_H4 <- predict(H4_ranger, newdata = dat_test)

predictions_H1H2 <- predict(H1H2_ranger, newdata = dat_test)
predictions_H1H3 <- predict(H1H3_ranger, newdata = dat_test)
predictions_H1H4 <- predict(H1H4_ranger, newdata = dat_test)
predictions_H2H3 <- predict(H2H3_ranger, newdata = dat_test)
predictions_H2H4 <- predict(H2H4_ranger, newdata = dat_test)
predictions_H3H4 <- predict(H3H4_ranger, newdata = dat_test)

predictions_H1H2H3 <- predict(H1H2H3_ranger, newdata = dat_test)
predictions_H1H2H4 <- predict(H1H2H4_ranger, newdata = dat_test)
predictions_H1H3H4 <- predict(H1H3H4_ranger, newdata = dat_test)
predictions_H2H3H4 <- predict(H2H3H4_ranger, newdata = dat_test)

# Calculate the performance metrics
perf <- rbind(postResample(pred = predictions_full, obs = dat_test$Jaccard),
              postResample(pred = predictions_H1, obs = dat_test$Jaccard),
              postResample(pred = predictions_H2, obs = dat_test$Jaccard),
              postResample(pred = predictions_H3, obs = dat_test$Jaccard),
              postResample(pred = predictions_H4, obs = dat_test$Jaccard),
              
              postResample(pred = predictions_H1H2, obs = dat_test$Jaccard),
              postResample(pred = predictions_H1H3, obs = dat_test$Jaccard),
              postResample(pred = predictions_H1H4, obs = dat_test$Jaccard),
              postResample(pred = predictions_H2H3, obs = dat_test$Jaccard),
              postResample(pred = predictions_H2H4, obs = dat_test$Jaccard),
              postResample(pred = predictions_H3H4, obs = dat_test$Jaccard),
              
              postResample(pred = predictions_H1H2H3, obs = dat_test$Jaccard),
              postResample(pred = predictions_H1H2H4, obs = dat_test$Jaccard),
              postResample(pred = predictions_H1H3H4, obs = dat_test$Jaccard),
              postResample(pred = predictions_H2H3H4, obs = dat_test$Jaccard)) %>% 
                as.data.frame() %>% 
                round(4)

model <- c("full", 
            seq(1:4), 
            "H1H2", "H1H3", "H1H4", "H2H3", "H2H4", "H3H4", 
            "H1H2H3", "H1H2H4", "H1H3H4", "H2H3H4")
perf$model <- model

# Print the performance metrics

perf %>% 
    kableExtra::kable()
perf %>% 
    kableExtra::kable() %>% 
    write.csv("../data/performance_varExpl_rf_J2.csv")

slice_min(perf, RMSE) %>% 
    slice_max(Rsquared)


# Create a bar plot of variance explained
ggplot(perf, aes(x = reorder(model, RMSE), y = Rsquared)) +
  geom_bar(stat = "identity") +
  labs(x = "Model", y = "Variance Explained") +
  theme_minimal()

```

## Log Ratio - Sampling period 1

```{r}
#| label: var-part-rf-LR1
#| eval: FALSE
dat_train <- dat_train_LR1

trainControl <- trainControl(
    method = "repeatedcv",
    number = 10,
    repeats = 3,
    savePredictions = "final",
    returnResamp = "final",
    verboseIter = TRUE,
    index = indices)

tictoc::tic("ranger full model")
set.seed(42)
full_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train,
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


# Train ranger model
tictoc::tic("ranger H1")
set.seed(42)
H1_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(H1_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


tictoc::tic("ranger H2")
set.seed(42)
H2_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(H2_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


tictoc::tic("ranger H3")
set.seed(42)
H3_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(H3_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()

tictoc::tic("ranger H4")
set.seed(42)
H4_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(H4_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


### combinations of 2 hypotheses:

set.seed(42)
H1H2_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H3_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H4_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)


set.seed(42)
H2H3_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H2_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)


set.seed(42)
H2H4_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H2_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H3H4_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H3_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)


### combinations of 3 hypotheses together =====

set.seed(42)
H1H2H3_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H2H4_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H3H4_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H3_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H2H3H4_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

LR1_varPart_ranger_list <- list(
    full_ranger, H1_ranger, H2_ranger, H3_ranger, H4_ranger, 
    H1H2_ranger, H1H3_ranger, H1H4_ranger, 
    H2H3_ranger, H2H4_ranger, 
    H3H4_ranger, 
    H1H2H3_ranger, H1H2H4_ranger, H1H3H4_ranger, 
    H2H3H4_ranger)
saveRDS(LR1_varPart_ranger_list, "../data/LR1_varPart_list.rds")

```

```{r}
#| label: var-part-ranger-performance-eval
# ===== Performance eval ======= #

# Predict on your test data
predictions_full <- predict(full_ranger, newdata = dat_test)

predictions_H1 <- predict(H1_ranger, newdata = dat_test)
predictions_H2 <- predict(H2_ranger, newdata = dat_test)
predictions_H3 <- predict(H3_ranger, newdata = dat_test)
predictions_H4 <- predict(H4_ranger, newdata = dat_test)

predictions_H1H2 <- predict(H1H2_ranger, newdata = dat_test)
predictions_H1H3 <- predict(H1H3_ranger, newdata = dat_test)
predictions_H1H4 <- predict(H1H4_ranger, newdata = dat_test)
predictions_H2H3 <- predict(H2H3_ranger, newdata = dat_test)
predictions_H2H4 <- predict(H2H4_ranger, newdata = dat_test)
predictions_H3H4 <- predict(H3H4_ranger, newdata = dat_test)

predictions_H1H2H3 <- predict(H1H2H3_ranger, newdata = dat_test)
predictions_H1H2H4 <- predict(H1H2H4_ranger, newdata = dat_test)
predictions_H1H3H4 <- predict(H1H3H4_ranger, newdata = dat_test)
predictions_H2H3H4 <- predict(H2H3H4_ranger, newdata = dat_test)

# Calculate the performance metrics
perf <- rbind(postResample(pred = predictions_full, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H1, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H2, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H3, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H4, obs = dat_test$log_R2_1),
              
              postResample(pred = predictions_H1H2, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H1H3, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H1H4, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H2H3, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H2H4, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H3H4, obs = dat_test$log_R2_1),
              
              postResample(pred = predictions_H1H2H3, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H1H2H4, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H1H3H4, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H2H3H4, obs = dat_test$log_R2_1)) %>% 
                as.data.frame() %>% 
                round(4)

model <- c("full", 
            seq(1:4), 
            "H1H2", "H1H3", "H1H4", "H2H3", "H2H4", "H3H4", 
            "H1H2H3", "H1H2H4", "H1H3H4", "H2H3H4")
perf$model <- model

# Print the performance metrics

perf %>% 
    kableExtra::kable()
perf %>% 
    kableExtra::kable() %>% 
    write.csv("../data/performance_varExpl_rf_LR1.csv")

slice_min(perf, RMSE) %>% 
    slice_max(Rsquared)


# Create a bar plot of variance explained
ggplot(perf, aes(x = reorder(model, RMSE), y = Rsquared)) +
  geom_bar(stat = "identity") +
  labs(x = "Model", y = "Variance Explained") +
  theme_minimal()

```

## Log Ratio - Sampling period 2

```{r}
#| label: var-part-rf-LR2
#| eval: FALSE
dat_train <- dat_train_LR2

trainControl <- trainControl(
    method = "repeatedcv",
    number = 10,
    repeats = 3,
    savePredictions = "final",
    returnResamp = "final",
    verboseIter = TRUE,
    index = indices)

tictoc::tic("ranger full model")
set.seed(42)
full_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train,
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


# Train ranger model
tictoc::tic("ranger H1")
set.seed(42)
H1_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(H1_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


tictoc::tic("ranger H2")
set.seed(42)
H2_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(H2_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


tictoc::tic("ranger H3")
set.seed(42)
H3_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(H3_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()

tictoc::tic("ranger H4")
set.seed(42)
H4_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(H4_vars)),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)
tictoc::toc()


### combinations of 2 hypotheses:

set.seed(42)
H1H2_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H3_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H4_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)


set.seed(42)
H2H3_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H2_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)


set.seed(42)
H2H4_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H2_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H3H4_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H3_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)


### combinations of 3 hypotheses together =====

set.seed(42)
H1H2H3_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H2H4_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H1H3H4_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H3_vars, H4_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

set.seed(42)
H2H3H4_ranger <- train(
    log_R2_1 ~ .,
    data = dat_train %>% select(Jaccard, any_of(c(H1_vars, H2_vars, H3_vars))),
    method = "ranger",
    trControl = trainControl,
    importance = "permutation",
    scale.permutation.importance = TRUE,
    num.trees = 5000,
    respect.unordered.factors = TRUE,
    oob.error = TRUE,
    tuneLength = 5)

LR2_varPart_ranger_list <- list(
    full_ranger, H1_ranger, H2_ranger, H3_ranger, H4_ranger, 
    H1H2_ranger, H1H3_ranger, H1H4_ranger, 
    H2H3_ranger, H2H4_ranger, 
    H3H4_ranger, 
    H1H2H3_ranger, H1H2H4_ranger, H1H3H4_ranger, 
    H2H3H4_ranger)
saveRDS(LR2_varPart_ranger_list, "../data/LR2_varPart_list.rds")

```

```{r}
#| label: var-part-ranger-performance-eval
# ===== Performance eval ======= #

# Predict on your test data
predictions_full <- predict(full_ranger, newdata = dat_test)

predictions_H1 <- predict(H1_ranger, newdata = dat_test)
predictions_H2 <- predict(H2_ranger, newdata = dat_test)
predictions_H3 <- predict(H3_ranger, newdata = dat_test)
predictions_H4 <- predict(H4_ranger, newdata = dat_test)

predictions_H1H2 <- predict(H1H2_ranger, newdata = dat_test)
predictions_H1H3 <- predict(H1H3_ranger, newdata = dat_test)
predictions_H1H4 <- predict(H1H4_ranger, newdata = dat_test)
predictions_H2H3 <- predict(H2H3_ranger, newdata = dat_test)
predictions_H2H4 <- predict(H2H4_ranger, newdata = dat_test)
predictions_H3H4 <- predict(H3H4_ranger, newdata = dat_test)

predictions_H1H2H3 <- predict(H1H2H3_ranger, newdata = dat_test)
predictions_H1H2H4 <- predict(H1H2H4_ranger, newdata = dat_test)
predictions_H1H3H4 <- predict(H1H3H4_ranger, newdata = dat_test)
predictions_H2H3H4 <- predict(H2H3H4_ranger, newdata = dat_test)

# Calculate the performance metrics
perf <- rbind(postResample(pred = predictions_full, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H1, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H2, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H3, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H4, obs = dat_test$log_R2_1),
              
              postResample(pred = predictions_H1H2, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H1H3, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H1H4, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H2H3, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H2H4, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H3H4, obs = dat_test$log_R2_1),
              
              postResample(pred = predictions_H1H2H3, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H1H2H4, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H1H3H4, obs = dat_test$log_R2_1),
              postResample(pred = predictions_H2H3H4, obs = dat_test$log_R2_1)) %>% 
                as.data.frame() %>% 
                round(4)

model <- c("full", 
            seq(1:4), 
            "H1H2", "H1H3", "H1H4", "H2H3", "H2H4", "H3H4", 
            "H1H2H3", "H1H2H4", "H1H3H4", "H2H3H4")
perf$model <- model

# Print the performance metrics

perf %>% 
    kableExtra::kable()
perf %>% 
    kableExtra::kable() %>% 
    write.csv("../data/performance_varExpl_rf_LR2.csv")

slice_min(perf, RMSE) %>% 
    slice_max(Rsquared)


# Create a bar plot of variance explained
ggplot(perf, aes(x = reorder(model, RMSE), y = Rsquared)) +
  geom_bar(stat = "identity") +
  labs(x = "Model", y = "Variance Explained") +
  theme_minimal()

```
:::

```{r}
save.image("../data/RData/04_VarPart.RData")
```
